---
title: "Local RNA-seq QC Report"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    code_folding: hide
params:
  results_dir: "./results"
  report_title: "Local RNA-seq QC Report"
  run_manifest: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages({
  library(data.table)
  library(knitr)
  library(DT)
})

results_dir <- params$results_dir
run_manifest <- params$run_manifest

module_enabled <- function(module_name) {
  if (is.null(run_manifest) || run_manifest == "" || !file.exists(run_manifest)) {
    return(TRUE)
  }

  manifest <- data.table::fread(run_manifest)

  if (!all(c("module", "enabled") %in% names(manifest))) {
    return(TRUE)
  }

  match_row <- manifest[manifest$module == module_name]

  if (nrow(match_row) != 1) {
    return(TRUE)
  }

  isTRUE(match_row$enabled[[1]])
}

plot_path <- function(...) {
  file.path(results_dir, ...)
}

include_png_if_exists <- function(path) {
  if (file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    cat(paste0("*Missing plot:* `", path, "`\n"))
  }
}

show_table_if_exists <- function(path, caption = NULL) {
  if (file.exists(path)) {
    dt <- data.table::fread(path)
    
    if (!is.null(caption)) {
      cat(paste0("**", caption, "**\n\n"))
    }
    
    DT::datatable(
      dt,
      options = list(
        pageLength = min(10, nrow(dt)),
        scrollX = TRUE
      ),
      rownames = FALSE
    )
  } else {
    cat(paste0("*Missing table:* `", path, "`\n"))
  }
}
```

# `r params$report_title`

Generated on: `r Sys.time()`

This report summarizes the local RNA-seq/cfRNA QC outputs generated by the pipeline.

---

# 1. Run status

```{r run-status}
status_path <- plot_path("run_logs", "local_qc_run_status.tsv")
show_table_if_exists(status_path, "Pipeline step status")
```

---

# 2. Normalization and log-transformation QC

This section checks whether the raw HTSeq counts were successfully normalized and log-transformed.

## Before/after histogram

```{r norm-hist, fig.align='center', out.width='100%'}
include_png_if_exists(plot_path("norm_log", "qc_before_after_hist.png"))
```

## Library size check

```{r libsize, fig.align='center', out.width='80%'}
include_png_if_exists(plot_path("norm_log", "qc_librarysize_raw_vs_norm.png"))
```

## Density curves before normalization

```{r density-before-nonzero, fig.align='center', out.width='100%'}
include_png_if_exists(plot_path("norm_log", "qc_density_before_nonzero.png"))
```

## Density curves after normalization

```{r density-after-nonzero, fig.align='center', out.width='100%'}
include_png_if_exists(plot_path("norm_log", "qc_density_after_nonzero.png"))
```

---

```{r biotype-heading, results='asis'}
if (module_enabled("biotype")) {
  cat("---\n\n# 3. Biotype distribution QC\n\n")
  cat("This section shows the fraction of raw counts assigned to different gene biotypes.\n")
}
```

```{r biotype-plot, results='asis', echo=FALSE}
if (module_enabled("biotype")) {
  full_biotype_plot <- plot_path("biotype_qc", "biotype_distribution_percent.png")
  chunk_dir <- plot_path("biotype_qc", "biotype_distribution_chunks")

  if (file.exists(full_biotype_plot)) {
    cat("\n\n## Biotype distribution plot\n\n")
    cat(paste0("![](", full_biotype_plot, ")\n\n"))
  } else if (dir.exists(chunk_dir)) {
    chunk_plots <- list.files(
      chunk_dir,
      pattern = "^biotype_distribution_percent_part_.*\\.png$",
      full.names = TRUE
    )

    if (length(chunk_plots) > 0) {
      chunk_plots <- sort(chunk_plots)

      cat("\n\n## Biotype distribution plots by sample chunks\n\n")

      for (plot_file in chunk_plots) {
        plot_name <- basename(plot_file)
        part_label <- gsub("^biotype_distribution_percent_", "", plot_name)
        part_label <- gsub("\\.png$", "", part_label)
        part_label <- gsub("_", " ", part_label)

        cat(paste0("\n\n### ", part_label, "\n\n"))
        cat(paste0("![](", plot_file, ")\n\n"))
      }
    } else {
      cat("*No biotype distribution plots were found.*\n")
    }
  } else {
    cat(paste0("*Missing plot:* `", full_biotype_plot, "`\n"))
  }
}
```

```{r biotype-table}
if (module_enabled("biotype")) {
  show_table_if_exists(
    plot_path("biotype_qc", "biotype_distribution_counts_and_fraction.tsv"),
    "Biotype distribution table"
  )
}
```

---

```{r pca-heading, results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("# 4. PCA on top variable genes\n\n")
  cat("The PCA is calculated from the log-normalized expression matrix using the top variable genes.\n")
}
```

```{r pca-scatter, fig.align='center', out.width='90%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("## PCA scatter plot\n\n")
  include_png_if_exists(plot_path("pca", "pca_scatter_PC1_PC2.png"))
}
```

```{r pca-scree, fig.align='center', out.width='80%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("## Scree plot\n\n")
  include_png_if_exists(plot_path("pca", "pca_scree.png"))
}
```

```{r pca-metadata-plots, results='asis', echo=FALSE}
if (module_enabled("pca_ssgsea")) {
  metadata_plot_dir <- plot_path("pca", "metadata_plots")

  if (dir.exists(metadata_plot_dir)) {
    metadata_plots <- list.files(
      metadata_plot_dir,
      pattern = "^pca_scatter_PC1_PC2_by_.*\\.png$",
      full.names = TRUE
    )
    
    if (length(metadata_plots) > 0) {
      metadata_plots <- sort(metadata_plots)
      cat("## PCA plots colored by sample metadata\n\n")

      for (plot_file in metadata_plots) {
        plot_name <- basename(plot_file)
        plot_label <- gsub("^pca_scatter_PC1_PC2_by_", "", plot_name)
        plot_label <- gsub("\\.png$", "", plot_label)
        plot_label <- gsub("_", " ", plot_label)

        cat(paste0("\n\n### Colored by ", plot_label, "\n\n"))
        cat(paste0("![](", plot_file, ")\n\n"))
      }
    } else {
      cat("*No metadata-colored PCA plots were generated.*\n")
    }
  } else {
    cat("*No metadata-colored PCA plots were generated.*\n")
  }
}
```

```{r pca-var-table, results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("## PCA variance explained\n\n")
  show_table_if_exists(
    plot_path("pca", "pca_variance_explained.tsv"),
    "PCA variance explained"
  )
}
```

---

```{r ssgsea-heading, results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("# 5. Gene-set contamination scores\n\n")
  cat("This section reports ssGSEA scores for platelet, erythroid/hemolysis, and endothelial marker sets.\n")
}
```

```{r ssgsea-pc-heatmap, fig.align='center', out.width='100%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("## ssGSEA score versus PC correlation heatmap\n\n")
  include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "ssgsea_score_pc_correlation_heatmap.png"))
}
```

```{r platelet-pc1, fig.align='center', out.width='80%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("## Platelet score vs PC1\n\n")
  include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "platelet_score_vs_PC1_ssgsea.png"))
}
```

```{r erythroid-pc1, fig.align='center', out.width='80%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("## Erythroid/hemolysis score vs PC1\n\n")
  include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "erythroid_score_vs_PC1_ssgsea.png"))
}
```

```{r endothelial-pc1, fig.align='center', out.width='80%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("## Endothelial score vs PC1\n\n")
  include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "endothelial_score_vs_PC1_ssgsea.png"))
}
```

```{r ssgsea-score-table, results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("## ssGSEA scores table\n\n")
  show_table_if_exists(
    plot_path("gene_set_qc", "ssgsea", "ssgsea_scores.tsv"),
    "ssGSEA scores"
  )
}
```

```{r ssgsea-marker-stats, results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("## Marker matching statistics\n\n")
  show_table_if_exists(
    plot_path("gene_set_qc", "ssgsea", "ssgsea_marker_match_stats.tsv"),
    "Marker matching statistics"
  )
}
```

---

```{r sex-heading, results='asis'}
if (module_enabled("sex_inference")) {
  cat("# 6. Sex chromosome QC\n\n")
  cat("This section visualizes XIST expression versus the mean expression of a small Y-linked gene panel.\n")
}
```

```{r sex-scatter, fig.align='center', out.width='80%'}
if (module_enabled("sex_inference")) {
  include_png_if_exists(plot_path("sex_inference", "sex_inference_scatter.png"))
}
```

```{r sex-scores}
if (module_enabled("sex_inference")) {
  show_table_if_exists(
    plot_path("sex_inference", "sex_inference_scores.tsv"),
    "Sex chromosome QC scores"
  )
}
```

```{r sex-marker-stats}
if (module_enabled("sex_inference")) {
  show_table_if_exists(
    plot_path("sex_inference", "sex_inference_marker_match_stats.tsv"),
    "Sex chromosome marker matching statistics"
  )
}
```

---

```{r hpc-heading, results='asis'}
if (module_enabled("hpc_metric")) {
  cat("# 7. HPC QC metric correlation with expression PCs\n\n")
  cat("This section compares upstream HPC-derived QC metrics against expression PCs.\n")
}
```

```{r hpc-pc-heatmap, fig.align='center', out.width='100%', results='asis'}
if (module_enabled("hpc_metric")) {
  cat("## HPC QC metric-PC correlation heatmap\n\n")
  include_png_if_exists(plot_path("hpc_metric_pc_qc", "hpc_qc_pc_correlation_heatmap.png"))
}
```

```{r hpc-cor-table, results='asis'}
if (module_enabled("hpc_metric")) {
  show_table_if_exists(
    plot_path("hpc_metric_pc_qc", "hpc_qc_pc_correlations.tsv"),
    "HPC QC metric-PC correlations"
  )
}
```

```{r hpc-z-table, results='asis'}
if (module_enabled("hpc_metric")) {
  cat("## Z-scored HPC QC metrics\n\n")
  show_table_if_exists(
    plot_path("hpc_metric_pc_qc", "hpc_qc_metrics_zscore.tsv"),
    "Z-scored HPC QC metrics"
  )
}
```

---

# 9. Notes

- Correlation heatmaps show **Pearson r** values.
- Significance stars, when present, use BH-adjusted p-values:
  - `*` adjusted p < 0.05
  - `**` adjusted p < 0.01
  - `***` adjusted p < 0.001
- Missing optional plots/tables usually mean that the corresponding optional input was not provided, or that the module was skipped.
