---
title: "Local RNA-seq QC Report"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    code_folding: hide
params:
  results_dir: "./results"
  report_title: "Local RNA-seq QC Report"
  run_manifest: ""
---

<style>
details {
  margin: 1rem 0;
}

details > summary {
  cursor: pointer;
  display: inline-block;
  padding: 0.4rem 0.75rem;
  border: 1px solid #337ab7;
  border-radius: 4px;
  background: #f5f5f5;
  color: #337ab7;
  font-weight: 600;
}

details[open] > summary {
  margin-bottom: 0.75rem;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages({
  library(data.table)
  library(knitr)
  library(DT)
})

results_dir <- params$results_dir
run_manifest <- params$run_manifest

module_enabled <- function(module_name) {
  if (is.null(run_manifest) || run_manifest == "" || !file.exists(run_manifest)) {
    return(TRUE)
  }

  manifest <- data.table::fread(run_manifest)

  if (!all(c("module", "enabled") %in% names(manifest))) {
    return(TRUE)
  }

  match_row <- manifest[manifest$module == module_name]

  if (nrow(match_row) != 1) {
    return(TRUE)
  }

  isTRUE(match_row$enabled[[1]])
}

plot_path <- function(...) {
  file.path(results_dir, ...)
}

include_png_if_exists <- function(path) {
  if (file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    cat(paste0("*Missing plot:* `", path, "`\n"))
  }
}

show_table_if_exists <- function(path, caption = NULL, show_label = NULL) {
  if (file.exists(path)) {
    dt <- data.table::fread(path)

    table_widget <- DT::datatable(
      dt,
      options = list(
        pageLength = min(10, nrow(dt)),
        scrollX = TRUE
      ),
      rownames = FALSE
    )

    if (is.null(show_label)) {
      if (!is.null(caption)) {
        cat(paste0("**", caption, "**\n\n"))
      }
      return(table_widget)
    }

    details_contents <- list(htmltools::tags$summary(show_label))

    if (!is.null(caption)) {
      details_contents <- c(
        details_contents,
        list(htmltools::tags$p(htmltools::tags$strong(caption)))
      )
    }

    details_contents <- c(details_contents, list(table_widget))
    return(do.call(htmltools::tags$details, details_contents))
  } else {
    cat(paste0("*Missing table:* `", path, "`\n"))
  }
}

show_run_status_summary <- function(path) {
  if (!file.exists(path)) {
    cat(paste0("*Missing table:* `", path, "`\n"))
    return(invisible(NULL))
  }

  status_dt <- data.table::fread(path)

  if (!"status" %in% names(status_dt)) {
    cat("*Run-status table does not contain a status column.*\n")
    return(invisible(NULL))
  }

  unsuccessful <- status_dt$status %in% c("FAIL", "MISSING")

  if (!any(unsuccessful)) {
    cat("All scheduled modules ran correctly.\n")
  } else {
    cat("One or more modules did not complete successfully. Open the run-status table for details.\n")
  }
}
```

# `r params$report_title`

Generated on: `r Sys.time()`

This report summarizes the local RNA-seq/cfRNA QC outputs generated by the pipeline.

---

# 1. Run status

```{r run-status}
status_path <- plot_path("run_logs", "local_qc_run_status.tsv")
show_run_status_summary(status_path)
```

```{r run-status-table}
show_table_if_exists(
  status_path,
  "Pipeline step status",
  "Show run-status table"
)
```

---

# 2. Normalization and log-transformation QC

DESeq2 size-factor normalization adjusts counts for differences in sequencing
depth and library composition, while log transformation stabilizes the
mean-variance relationship. After normalization, sample density curves should
generally become more comparable, reflecting reduced technical scale
differences. Marked residual separation may still reflect genuine biological
heterogeneity or remaining technical variation and should be interpreted with
the study design in mind.

<details>
<summary>Show additional normalization plots</summary>

<h2>Before/after histogram</h2>

```{r norm-hist, fig.align='center', out.width='100%'}
include_png_if_exists(plot_path("norm_log", "qc_before_after_hist.png"))
```

<h2>Library size check</h2>

```{r libsize, fig.align='center', out.width='80%'}
include_png_if_exists(plot_path("norm_log", "qc_librarysize_raw_vs_norm.png"))
```

</details>

## Density curves before normalization

```{r density-before-nonzero, fig.align='center', out.width='100%'}
include_png_if_exists(plot_path("norm_log", "qc_density_before_nonzero.png"))
```

## Density curves after normalization

```{r density-after-nonzero, fig.align='center', out.width='100%'}
include_png_if_exists(plot_path("norm_log", "qc_density_after_nonzero.png"))
```

---

```{r biotype-heading, results='asis'}
if (module_enabled("biotype")) {
  cat("---\n\n# 3. Biotype distribution QC\n\n")
  cat(
    "This count-level QC assesses whether reads are predominantly assigned to expected transcript classes, such as protein-coding genes and long non-coding RNAs (lncRNAs). A typical RNA-seq profile is expected to be enriched for these biologically plausible classes, whereas elevated fractions of unexpected classes should prompt review of library preparation, annotation compatibility, alignment, and count assignment.\n"
  )
}
```

```{r biotype-plot, results='asis', echo=FALSE}
if (module_enabled("biotype")) {
  full_biotype_plot <- plot_path("biotype_qc", "biotype_distribution_percent.png")
  chunk_dir <- plot_path("biotype_qc", "biotype_distribution_chunks")

  if (file.exists(full_biotype_plot)) {
    cat("\n\n## Biotype distribution plot\n\n")
    cat(paste0("![](", full_biotype_plot, ")\n\n"))
  } else if (dir.exists(chunk_dir)) {
    chunk_plots <- list.files(
      chunk_dir,
      pattern = "^biotype_distribution_percent_part_.*\\.png$",
      full.names = TRUE
    )

    if (length(chunk_plots) > 0) {
      chunk_plots <- sort(chunk_plots)

      cat("\n\n## Biotype distribution plots by sample chunks\n\n")

      for (plot_file in chunk_plots) {
        plot_name <- basename(plot_file)
        part_label <- gsub("^biotype_distribution_percent_", "", plot_name)
        part_label <- gsub("\\.png$", "", part_label)
        part_label <- gsub("_", " ", part_label)

        cat(paste0("\n\n### ", part_label, "\n\n"))
        cat(paste0("![](", plot_file, ")\n\n"))
      }
    } else {
      cat("*No biotype distribution plots were found.*\n")
    }
  } else {
    cat(paste0("*Missing plot:* `", full_biotype_plot, "`\n"))
  }
}
```

```{r biotype-table, results='asis'}
if (module_enabled("biotype")) {
  show_table_if_exists(
    plot_path("biotype_qc", "biotype_distribution_counts_and_fraction.tsv"),
    "Biotype distribution table",
    "Show biotype distribution table"
  )
}
```

---

```{r pca-heading, results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("# 4. PCA on top variable genes\n\n")
  cat(
    "Principal component analysis (PCA) is a dimensionality-reduction method that summarizes the high-dimensional expression matrix into orthogonal principal components (PCs), ordered by the amount of variation they explain. Samples that lie close together in PCA space have more similar overall expression profiles.\n\nThe scree plot shows the proportion of variability captured by successive PCs. When metadata are supplied, the coloured PCA plots help assess whether sample separation is aligned with the experimental condition, as may be expected for a biological contrast, or with batch and other technical annotations, which can indicate unwanted variation.\n"
  )
}
```

```{r pca-scatter, fig.align='center', out.width='90%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("## PCA scatter plot\n\n")
  include_png_if_exists(plot_path("pca", "pca_scatter_PC1_PC2.png"))
}
```

```{r pca-scree, fig.align='center', out.width='80%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("## Scree plot\n\n")
  include_png_if_exists(plot_path("pca", "pca_scree.png"))
}
```

```{r pca-metadata-plots, results='asis', echo=FALSE}
if (module_enabled("pca_ssgsea")) {
  metadata_plot_dir <- plot_path("pca", "metadata_plots")

  if (dir.exists(metadata_plot_dir)) {
    metadata_plots <- list.files(
      metadata_plot_dir,
      pattern = "^pca_scatter_PC1_PC2_by_.*\\.png$",
      full.names = TRUE
    )
    
    if (length(metadata_plots) > 0) {
      metadata_plots <- sort(metadata_plots)
      cat("## PCA plots colored by sample metadata\n\n")

      for (plot_file in metadata_plots) {
        plot_name <- basename(plot_file)
        plot_label <- gsub("^pca_scatter_PC1_PC2_by_", "", plot_name)
        plot_label <- gsub("\\.png$", "", plot_label)
        plot_label <- gsub("_", " ", plot_label)

        cat(paste0("\n\n### Colored by ", plot_label, "\n\n"))
        cat(paste0("![](", plot_file, ")\n\n"))
      }
    } else {
      cat("*No metadata-colored PCA plots were generated.*\n")
    }
  } else {
    cat("*No metadata-colored PCA plots were generated.*\n")
  }
}
```

```{r pca-var-table, results='asis'}
if (module_enabled("pca_ssgsea")) {
  show_table_if_exists(
    plot_path("pca", "pca_variance_explained.tsv"),
    "PCA variance explained",
    "Show PCA variance-explained table"
  )
}
```

---

```{r ssgsea-heading, results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("# 5. Gene-set contamination scores\n\n")
  cat(
    "Single-sample gene-set enrichment analysis (ssGSEA) produces a per-sample enrichment score for each marker set. The platelet, erythroid/hemolysis, and endothelial sets are included because these signals can represent problematic contamination or compositional effects in cfRNA data. The heatmap tests whether these scores track the major expression PCs; the absence of strong or statistically supported score-PC correlations is generally reassuring, whereas a substantial significant association suggests that a leading PC may reflect one of these contamination-associated signals.\n"
  )
}
```

```{r ssgsea-pc-heatmap, fig.align='center', out.width='100%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("## ssGSEA score versus PC correlation heatmap\n\n")
  include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "ssgsea_score_pc_correlation_heatmap.png"))
}
```

```{r platelet-pc1, fig.align='center', out.width='80%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("<details><summary>Show individual gene-set plots and tables</summary>\n\n")
  cat("<h2>Platelet score vs PC1</h2>\n")
  include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "platelet_score_vs_PC1_ssgsea.png"))
}
```

```{r erythroid-pc1, fig.align='center', out.width='80%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("<h2>Erythroid/hemolysis score vs PC1</h2>\n")
  include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "erythroid_score_vs_PC1_ssgsea.png"))
}
```

```{r endothelial-pc1, fig.align='center', out.width='80%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("<h2>Endothelial score vs PC1</h2>\n")
  include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "endothelial_score_vs_PC1_ssgsea.png"))
  cat("\n</details>\n")
}
```

```{r ssgsea-score-table, results='asis'}
if (module_enabled("pca_ssgsea")) {
  show_table_if_exists(
    plot_path("gene_set_qc", "ssgsea", "ssgsea_scores.tsv"),
    "ssGSEA scores",
    "Show ssGSEA scores table"
  )
}
```

```{r ssgsea-marker-stats, results='asis'}
if (module_enabled("pca_ssgsea")) {
  show_table_if_exists(
    plot_path("gene_set_qc", "ssgsea", "ssgsea_marker_match_stats.tsv"),
    "Marker matching statistics",
    "Show ssGSEA marker matching statistics"
  )
}
```

---

```{r sex-heading, results='asis'}
if (module_enabled("sex_inference")) {
  cat("# 6. Sex chromosome QC\n\n")
  cat(
    "XIST is a long non-coding RNA expressed from the inactive X chromosome and is therefore typically detected in female samples. The y-axis shows the mean log-normalized expression of selected Y-chromosome genes: EIF1AY, KDM5D, UTY, DDX3Y, and RPS4Y1. A typical male pattern is low or absent XIST expression with higher Y-gene expression, with the reciprocal pattern expected for female samples; deviations should be interpreted alongside sample metadata and the known biology of the cohort.\n"
  )
}
```

```{r sex-scatter, fig.align='center', out.width='80%'}
if (module_enabled("sex_inference")) {
  include_png_if_exists(plot_path("sex_inference", "sex_inference_scatter.png"))
}
```

```{r sex-scores, results='asis'}
if (module_enabled("sex_inference")) {
  show_table_if_exists(
    plot_path("sex_inference", "sex_inference_scores.tsv"),
    "Sex chromosome QC scores",
    "Show sex chromosome QC scores table"
  )
}
```

```{r sex-marker-stats, results='asis'}
if (module_enabled("sex_inference")) {
  show_table_if_exists(
    plot_path("sex_inference", "sex_inference_marker_match_stats.tsv"),
    "Sex chromosome marker matching statistics",
    "Show sex chromosome marker matching-statistics table"
  )
}
```

---

```{r hpc-heading, results='asis'}
if (module_enabled("hpc_metric")) {
  cat("# 7. HPC QC metric correlation with expression PCs\n\n")
  cat(
    "This analysis evaluates whether upstream HPC-derived QC metrics are associated with the major expression PCs, thereby helping to identify potential sources of technical variation. As for the gene-set analysis, strong statistically supported correlations suggest that a PC may partly reflect a technical QC characteristic rather than the biological structure of interest.\n"
  )
}
```

```{r hpc-pc-heatmap, fig.align='center', out.width='100%', results='asis'}
if (module_enabled("hpc_metric")) {
  cat("## HPC QC metric-PC correlation heatmap\n\n")
  include_png_if_exists(plot_path("hpc_metric_pc_qc", "hpc_qc_pc_correlation_heatmap.png"))
}
```

```{r hpc-cor-table, results='asis'}
if (module_enabled("hpc_metric")) {
  show_table_if_exists(
    plot_path("hpc_metric_pc_qc", "hpc_qc_pc_correlations.tsv"),
    "HPC QC metric-PC correlations",
    "Show HPC QC metric-PC correlations table"
  )
}
```

```{r hpc-z-table, results='asis'}
if (module_enabled("hpc_metric")) {
  show_table_if_exists(
    plot_path("hpc_metric_pc_qc", "hpc_qc_metrics_zscore.tsv"),
    "Z-scored HPC QC metrics",
    "Show z-scored HPC QC metrics table"
  )
}
```

---

# Notes

- Correlation heatmaps show **Pearson r** values.
- Significance stars, when present, use BH-adjusted p-values:
  - `*` adjusted p < 0.05
  - `**` adjusted p < 0.01
  - `***` adjusted p < 0.001
- Missing optional plots/tables usually mean that the corresponding optional input was not provided, or that the module was skipped.
