---
title: "Local RNA-seq QC Report"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    code_folding: hide
params:
  results_dir: "./results"
  report_title: "Local RNA-seq QC Report"
  run_manifest: ""
---

<style>
details {
  margin: 1rem 0;
}

details > summary {
  cursor: pointer;
  display: inline-block;
  padding: 0.4rem 0.75rem;
  border: 1px solid #337ab7;
  border-radius: 4px;
  background: #f5f5f5;
  color: #337ab7;
  font-weight: 600;
}

details[open] > summary {
  margin-bottom: 0.75rem;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages({
  library(data.table)
  library(knitr)
  library(DT)
  library(plotly)
})

results_dir <- params$results_dir
run_manifest <- params$run_manifest

module_enabled <- function(module_name) {
  if (is.null(run_manifest) || run_manifest == "" || !file.exists(run_manifest)) {
    return(TRUE)
  }

  manifest <- data.table::fread(run_manifest)

  if (!all(c("module", "enabled") %in% names(manifest))) {
    return(TRUE)
  }

  match_row <- manifest[manifest$module == module_name]

  if (nrow(match_row) != 1) {
    return(TRUE)
  }

  isTRUE(match_row$enabled[[1]])
}

plot_path <- function(...) {
  file.path(results_dir, ...)
}

include_png_if_exists <- function(path) {
  if (file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    cat(paste0("*Missing plot:* `", path, "`\n"))
  }
}

show_interactive_or_png <- function(data_path, png_path, make_plot) {
  if (file.exists(data_path)) {
    make_plot(data.table::fread(data_path))
  } else {
    include_png_if_exists(png_path)
  }
}

interactive_scatter <- function(dt, sample_col, x_col, y_col, x_title, y_title,
                                color_col = NULL, plot_title = NULL) {
  hover <- paste0(
    "sample: ", htmltools::htmlEscape(as.character(dt[[sample_col]])),
    "<br>", x_title, ": ", signif(dt[[x_col]], 4),
    "<br>", y_title, ": ", signif(dt[[y_col]], 4)
  )
  plot_args <- list(
    x = dt[[x_col]], y = dt[[y_col]], type = "scatter", mode = "markers",
    text = hover, hoverinfo = "text", marker = list(size = 8), height = 470
  )
  if (!is.null(color_col)) {
    plot_args$color <- dt[[color_col]]
  }
  p <- do.call(plotly::plot_ly, plot_args)
  plotly::layout(
    p,
    title = if (is.null(plot_title)) NULL else list(text = plot_title, x = 0.5),
    xaxis = list(
      title = list(text = x_title, standoff = 12),
      showline = TRUE, linecolor = "#333333", linewidth = 1,
      ticks = "outside", showticklabels = TRUE, automargin = TRUE
    ),
    yaxis = list(
      title = list(text = y_title, standoff = 12),
      showline = TRUE, linecolor = "#333333", linewidth = 1,
      ticks = "outside", showticklabels = TRUE, automargin = TRUE
    ),
    margin = list(l = 95, b = 85, r = 30, t = if (is.null(plot_title)) 35 else 70),
    hovermode = "closest"
  )
}

interactive_density <- function(dt, x_title) {
  p <- plotly::plot_ly()
  for (sample_name in unique(dt$sample)) {
    curve <- dt[dt$sample == sample_name, ]
    hover <- paste0(
      "sample: ", htmltools::htmlEscape(sample_name),
      "<br>", x_title, ": ", signif(curve$value, 3),
      "<br>density: ", signif(curve$density, 3)
    )
    p <- plotly::add_lines(
      p, x = curve$value, y = curve$density, name = sample_name,
      text = hover, hoverinfo = "text", showlegend = FALSE
    )
  }
  plotly::layout(
    p,
    xaxis = list(title = x_title), yaxis = list(title = "density"),
    hovermode = "closest"
  )
}

pca_axis_titles <- function() {
  path <- plot_path("pca", "pca_variance_explained.tsv")
  if (!file.exists(path)) return(c("PC1", "PC2"))
  variance <- data.table::fread(path)
  v <- variance$variance_explained[match(c("PC1", "PC2"), variance$PC)]
  paste0(c("PC1", "PC2"), " (variance explained: ", round(100 * v, 1), "%)")
}

show_table_if_exists <- function(path, caption = NULL, show_label = NULL, transform = identity) {
  if (file.exists(path)) {
    dt <- transform(data.table::fread(path))

    table_widget <- DT::datatable(
      dt,
      options = list(
        pageLength = min(10, nrow(dt)),
        scrollX = TRUE
      ),
      rownames = FALSE
    )

    if (is.null(show_label)) {
      if (!is.null(caption)) {
        cat(paste0("**", caption, "**\n\n"))
      }
      return(table_widget)
    }

    details_contents <- list(htmltools::tags$summary(show_label))

    if (!is.null(caption)) {
      details_contents <- c(
        details_contents,
        list(htmltools::tags$p(htmltools::tags$strong(caption)))
      )
    }

    details_contents <- c(details_contents, list(table_widget))
    return(do.call(htmltools::tags$details, details_contents))
  } else {
    cat(paste0("*Missing table:* `", path, "`\n"))
  }
}

show_run_status_summary <- function(path) {
  if (!file.exists(path)) {
    cat(paste0("*Missing table:* `", path, "`\n"))
    return(invisible(NULL))
  }

  status_dt <- data.table::fread(path)

  if (!"status" %in% names(status_dt)) {
    cat("*Run-status table does not contain a status column.*\n")
    return(invisible(NULL))
  }

  unsuccessful <- status_dt$status %in% c("FAIL", "MISSING")

  if (!any(unsuccessful)) {
    cat("All scheduled modules ran correctly.\n")
  } else {
    cat("One or more modules did not complete successfully. Open the run-status table for details.\n")
  }
}
```

# `r params$report_title`

Generated on: `r Sys.time()`

This report summarizes the cfRNA QC outputs generated by the local tool.

---

# 1. Run status

```{r run-status}
status_path <- plot_path("run_logs", "local_qc_run_status.tsv")
show_run_status_summary(status_path)
```

```{r run-status-table}
show_table_if_exists(
  status_path,
  "Pipeline step status",
  "Show run-status table"
)
```

---

# 2. Normalization and log-transformation QC

DESeq2 size-factor normalization adjusts counts for differences in sequencing
depth and library composition, while log transformation stabilizes the
mean-variance relationship. After normalization, sample density curves should
generally become more comparable, reflecting reduced technical scale
differences. Marked residual separation may still reflect genuine biological
heterogeneity or remaining technical variation and should be interpreted with
the study design in mind.

## Library size check

```{r libsize, fig.align='center', out.width='80%'}
show_interactive_or_png(
  plot_path("norm_log", "qc_librarysize.tsv"),
  plot_path("norm_log", "qc_librarysize_raw_vs_norm.png"),
  function(dt) interactive_scatter(
    dt, "sample", "raw", "normalized", "raw library size",
    "normalized library size"
  )
)
```

## Density curves before normalization

```{r density-before-nonzero, fig.align='center', out.width='100%'}
show_interactive_or_png(
  plot_path("norm_log", "qc_density_before_nonzero.tsv"),
  plot_path("norm_log", "qc_density_before_nonzero.png"),
  function(dt) interactive_density(dt, "log2(raw counts + 1)")
)
```

## Density curves after normalization

```{r density-after-nonzero, fig.align='center', out.width='100%'}
show_interactive_or_png(
  plot_path("norm_log", "qc_density_after_nonzero.tsv"),
  plot_path("norm_log", "qc_density_after_nonzero.png"),
  function(dt) interactive_density(dt, "log2(normalized counts + 1)")
)
```

---

```{r biotype-heading, results='asis'}
if (module_enabled("biotype")) {
  cat("---\n\n# 3. Biotype distribution QC\n\n")
  cat(
    "This count-level QC shows how reads are distributed across transcript classes, such as protein-coding genes and long non-coding RNAs (lncRNAs). The expected distribution depends on the enrichment and library preparation protocol. A distribution that differs from your expectations, or large differences between samples prepared in the same way, should prompt review of library preparation, annotation compatibility, alignment, and count assignment.\n"
  )
}
```

```{r biotype-plot, results='asis', echo=FALSE}
if (module_enabled("biotype")) {
  full_biotype_plot <- plot_path("biotype_qc", "biotype_distribution_percent.png")
  chunk_dir <- plot_path("biotype_qc", "biotype_distribution_chunks")

  if (file.exists(full_biotype_plot)) {
    cat("\n\n## Biotype distribution plot\n\n")
    cat(paste0("![](", full_biotype_plot, ")\n\n"))
  } else if (dir.exists(chunk_dir)) {
    chunk_plots <- list.files(
      chunk_dir,
      pattern = "^biotype_distribution_percent_part_.*\\.png$",
      full.names = TRUE
    )

    if (length(chunk_plots) > 0) {
      chunk_plots <- sort(chunk_plots)

      cat("\n\n## Biotype distribution plots by sample chunks\n\n")

      for (plot_file in chunk_plots) {
        plot_name <- basename(plot_file)
        part_label <- gsub("^biotype_distribution_percent_", "", plot_name)
        part_label <- gsub("\\.png$", "", part_label)
        part_label <- gsub("_", " ", part_label)

        cat(paste0("\n\n### ", part_label, "\n\n"))
        cat(paste0("![](", plot_file, ")\n\n"))
      }
    } else {
      cat("*No biotype distribution plots were found.*\n")
    }
  } else {
    cat(paste0("*Missing plot:* `", full_biotype_plot, "`\n"))
  }
}
```

```{r biotype-table, results='asis'}
if (module_enabled("biotype")) {
  show_table_if_exists(
    plot_path("biotype_qc", "biotype_distribution_counts_and_fraction.tsv"),
    "Biotype distribution (% of counts per sample)",
    "Show biotype distribution table",
    transform = function(dt) {
      dt[, percent := round(100 * frac, 2)]
      data.table::dcast(dt, gene_biotype ~ sample_id, value.var = "percent")
    }
  )
}
```

---

```{r pca-heading, results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("# 4. PCA on top variable genes\n\n")
  cat(
    "Principal component analysis (PCA) is a dimensionality-reduction method that summarizes the high-dimensional expression matrix into orthogonal principal components (PCs), ordered by the amount of variation they explain. Samples that lie close together in PCA space have more similar overall expression profiles.\n\nThe scree plot shows the proportion of variability captured by successive PCs.\n\nWhen metadata are supplied, the coloured PCA plots help assess whether sample separation is aligned with the experimental condition, as may be expected for a biological contrast, or with batch and other technical annotations, which can indicate unwanted variation.\n"
  )
}
```

```{r pca-scatter, fig.align='center', out.width='90%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("## PCA scatter plot\n\n")
  axis_titles <- pca_axis_titles()
  show_interactive_or_png(
    plot_path("pca", "pca_scores.tsv"),
    plot_path("pca", "pca_scatter_PC1_PC2.png"),
    function(dt) interactive_scatter(
      dt, "sample_id", "PC1", "PC2", axis_titles[1], axis_titles[2]
    )
  )
}
```

```{r pca-scree, fig.align='center', out.width='80%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("## Scree plot\n\n")
  include_png_if_exists(plot_path("pca", "pca_scree.png"))
}
```

```{r pca-metadata-plots, results='asis', echo=FALSE}
if (module_enabled("pca_ssgsea")) {
  columns_path <- plot_path("pca", "pca_metadata_plot_columns.tsv")
  scores_path <- plot_path("pca", "pca_scores_with_metadata.tsv")
  if (file.exists(columns_path) && file.exists(scores_path)) {
    columns <- data.table::fread(columns_path)$metadata_column
    scores <- data.table::fread(scores_path)
    columns <- columns[columns %in% names(scores)]
    if (length(columns) > 0) {
      axis_titles <- pca_axis_titles()
      widgets <- lapply(columns, function(column) {
        htmltools::tagList(
          htmltools::tags$h3(paste("Colored by", column)),
          interactive_scatter(
            scores, "sample_id", "PC1", "PC2",
            axis_titles[1], axis_titles[2], color_col = column
          )
        )
      })
      htmltools::tagList(
        htmltools::tags$h2("PCA plots colored by sample metadata"),
        widgets
      )
    } else {
      cat("*No metadata-colored PCA plots were generated.*\n")
    }
  } else {
    metadata_plot_dir <- plot_path("pca", "metadata_plots")
    metadata_plots <- if (dir.exists(metadata_plot_dir)) {
      sort(list.files(
        metadata_plot_dir,
        pattern = "^pca_scatter_PC1_PC2_by_.*\\.png$",
        full.names = TRUE
      ))
    } else character()
    if (length(metadata_plots) > 0) {
      cat("## PCA plots colored by sample metadata\n\n")
      for (plot_file in metadata_plots) {
        plot_label <- sub("\\.png$", "", sub("^pca_scatter_PC1_PC2_by_", "", basename(plot_file)))
        cat(paste0("\n\n### Colored by ", gsub("_", " ", plot_label), "\n\n"))
        cat(paste0("![](", plot_file, ")\n\n"))
      }
    } else cat("*No metadata-colored PCA plots were generated.*\n")
  }
}
```

```{r pca-var-table, results='asis'}
if (module_enabled("pca_ssgsea")) {
  show_table_if_exists(
    plot_path("pca", "pca_variance_explained.tsv"),
    "PCA variance explained",
    "Show PCA variance-explained table"
  )
}
```

---

```{r ssgsea-heading, results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("# 5. Gene-set contamination scores\n\n")
  cat(
    "Single-sample gene-set enrichment analysis (ssGSEA) produces a per-sample enrichment score for each marker set. The platelet, erythroid, and endothelial sets are included because these signals can represent problematic contamination or compositional effects in cfRNA data. The heatmap tests whether these scores track the major expression PCs; the absence of strong or statistically supported score-PC correlations is generally reassuring, whereas a substantial significant association suggests that a leading PC may reflect one of these contamination-associated signals.\n"
  )
}
```

```{r ssgsea-pc-heatmap, fig.align='center', out.width='100%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("## ssGSEA score versus PC correlation heatmap\n\n")
  include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "ssgsea_score_pc_correlation_heatmap.png"))
}
```

```{r platelet-pc1, fig.align='center', out.width='80%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("<details><summary>Show individual gene-set plots</summary>\n\n")
  cat("<h2>Platelet score vs PC1</h2>\n")
  include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "platelet_score_vs_PC1_ssgsea.png"))
}
```

```{r erythroid-pc1, fig.align='center', out.width='80%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("<h2>Erythroid score vs PC1</h2>\n")
  include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "erythroid_score_vs_PC1_ssgsea.png"))
}
```

```{r endothelial-pc1, fig.align='center', out.width='80%', results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("<h2>Endothelial score vs PC1</h2>\n")
  include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "endothelial_score_vs_PC1_ssgsea.png"))
}
```

```{r close-gene-set-plots, results='asis'}
if (module_enabled("pca_ssgsea")) cat("</details>\n")
```

```{r ssgsea-score-table, results='asis'}
if (module_enabled("pca_ssgsea")) {
  show_table_if_exists(
    plot_path("gene_set_qc", "ssgsea", "ssgsea_scores.tsv"),
    "ssGSEA scores",
    "Show ssGSEA scores table"
  )
}
```

```{r ssgsea-marker-stats, results='asis'}
if (module_enabled("pca_ssgsea")) {
  cat("Marker matching shows how many genes in each reference marker set could be mapped to Ensembl IDs and found in this dataset. Sets with few matched genes have scores based on less marker coverage and should be interpreted cautiously.\n\n")
  show_table_if_exists(
    plot_path("gene_set_qc", "ssgsea", "ssgsea_marker_match_stats.tsv"),
    "ssGSEA marker matching overview",
    "Show ssGSEA marker matching overview"
  )
}
```

---

```{r sex-heading, results='asis'}
if (module_enabled("sex_inference")) {
  cat("# 6. Sex chromosome QC\n\n")
  cat(
    "*XIST* is a long non-coding RNA expressed from the inactive X chromosome and is therefore typically detected in female samples. The plot compares *XIST* expression with the mean log-normalized expression of five Y-chromosome genes: *EIF1AY*, *KDM5D*, *UTY*, *DDX3Y*, and *RPS4Y1*. These genes are commonly used in sex inference prediction models. A typical male pattern is low or absent *XIST* expression with higher Y-chromosome gene expression, with the reciprocal pattern expected for female samples; deviations should be interpreted alongside sample metadata and the known biology of the cohort.\n"
  )
}
```

```{r sex-scatter, fig.align='center', out.width='80%'}
if (module_enabled("sex_inference")) {
  show_interactive_or_png(
    plot_path("sex_inference", "sex_inference_scores.tsv"),
    plot_path("sex_inference", "sex_inference_scatter.png"),
    function(dt) interactive_scatter(
      dt, "sample_id", "Y_panel_mean_log2norm", "XIST_log2norm",
      "mean log2(normalized counts + 1) of Y-chromosome genes",
      "log2(normalized counts + 1) of <i>XIST</i>",
      plot_title = "sex chromosome QC: <i>XIST</i> vs mean Y-chromosome gene expression"
    )
  )
}
```

```{r sex-scores, results='asis'}
if (module_enabled("sex_inference")) {
  show_table_if_exists(
    plot_path("sex_inference", "sex_inference_scores.tsv"),
    "Sex chromosome QC scores",
    "Show sex chromosome QC scores table"
  )
}
```

```{r sex-marker-stats, results='asis'}
if (module_enabled("sex_inference")) {
  cat("This overview shows whether *XIST* and the five Y-chromosome genes were resolved to Ensembl IDs and found in the expression matrix. Missing markers reduce the evidence available for interpreting the sex chromosome QC plot.\n\n")
  show_table_if_exists(
    plot_path("sex_inference", "sex_inference_marker_match_stats.tsv"),
    "Sex chromosome marker matching overview",
    "Show sex chromosome marker matching overview"
  )
}
```

---

```{r hpc-heading, results='asis'}
if (module_enabled("hpc_metric")) {
  cat("# 7. HPC QC metric correlation with expression PCs\n\n")
  cat(
    "This analysis evaluates whether upstream HPC-derived QC metrics are associated with the major expression PCs, thereby helping to identify potential sources of technical variation. As for the gene-set analysis, strong statistically supported correlations suggest that a PC may partly reflect a technical QC characteristic rather than the biological structure of interest.\n"
  )
}
```

```{r hpc-pc-heatmap, fig.align='center', out.width='100%', results='asis'}
if (module_enabled("hpc_metric")) {
  cat("## HPC QC metric-PC correlation heatmap\n\n")
  include_png_if_exists(plot_path("hpc_metric_pc_qc", "hpc_qc_pc_correlation_heatmap.png"))
}
```

```{r hpc-cor-table, results='asis'}
if (module_enabled("hpc_metric")) {
  show_table_if_exists(
    plot_path("hpc_metric_pc_qc", "hpc_qc_pc_correlations.tsv"),
    "HPC QC metric-PC correlations",
    "Show HPC QC metric-PC correlations table"
  )
}
```

```{r hpc-z-table, results='asis'}
if (module_enabled("hpc_metric")) {
  show_table_if_exists(
    plot_path("hpc_metric_pc_qc", "hpc_qc_metrics_zscore.tsv"),
    "Z-scored HPC QC metrics",
    "Show z-scored HPC QC metrics table"
  )
}
```

---

# Notes

- Correlation heatmaps show **Pearson r** values.
- Significance stars, when present, use BH-adjusted p-values:
  - `*` adjusted p < 0.05
  - `**` adjusted p < 0.01
  - `***` adjusted p < 0.001
- Missing optional plots/tables usually mean that the corresponding optional input was not provided, or that the module was skipped.
