---
title: "Local RNA-seq QC Report"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    code_folding: hide
params:
  results_dir: "./results"
  report_title: "Local RNA-seq QC Report"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

suppressPackageStartupMessages({
  library(data.table)
  library(knitr)
  library(DT)
})

results_dir <- params$results_dir

plot_path <- function(...) {
  file.path(results_dir, ...)
}

include_png_if_exists <- function(path) {
  if (file.exists(path)) {
    knitr::include_graphics(path)
  } else {
    cat(paste0("*Missing plot:* `", path, "`\n"))
  }
}

show_table_if_exists <- function(path, caption = NULL) {
  if (file.exists(path)) {
    dt <- data.table::fread(path)
    
    if (!is.null(caption)) {
      cat(paste0("**", caption, "**\n\n"))
    }
    
    DT::datatable(
      dt,
      options = list(
        pageLength = min(10, nrow(dt)),
        scrollX = TRUE
      ),
      rownames = FALSE
    )
  } else {
    cat(paste0("*Missing table:* `", path, "`\n"))
  }
}
```

# `r params$report_title`

Generated on: `r Sys.time()`

This report summarizes the local RNA-seq/cfRNA QC outputs generated by the pipeline.

---

# 1. Run status

```{r run-status}
status_path <- plot_path("run_logs", "local_qc_run_status.tsv")
show_table_if_exists(status_path, "Pipeline step status")
```

---

# 2. Normalization and log-transformation QC

This section checks whether the raw HTSeq counts were successfully normalized and log-transformed.

## Before/after histogram

```{r norm-hist, fig.align='center', out.width='100%'}
include_png_if_exists(plot_path("norm_log", "qc_before_after_hist.png"))
```

## Library size check

```{r libsize, fig.align='center', out.width='80%'}
include_png_if_exists(plot_path("norm_log", "qc_librarysize_raw_vs_norm.png"))
```

## Density curves before normalization

```{r density-before-nonzero, fig.align='center', out.width='100%'}
include_png_if_exists(plot_path("norm_log", "qc_density_before_nonzero.png"))
```

## Density curves after normalization

```{r density-after-nonzero, fig.align='center', out.width='100%'}
include_png_if_exists(plot_path("norm_log", "qc_density_after_nonzero.png"))
```

---

# 3. Biotype distribution QC

This section shows the fraction of raw counts assigned to different gene biotypes.

```{r biotype-plot, results='asis', echo=FALSE}
full_biotype_plot <- plot_path("biotype_qc", "biotype_distribution_percent.png")
chunk_dir <- plot_path("biotype_qc", "biotype_distribution_chunks")

if (file.exists(full_biotype_plot)) {
  cat("\n\n## Biotype distribution plot\n\n")
  cat(paste0("![](", full_biotype_plot, ")\n\n"))
} else if (dir.exists(chunk_dir)) {
  chunk_plots <- list.files(
    chunk_dir,
    pattern = "^biotype_distribution_percent_part_.*\\.png$",
    full.names = TRUE
  )
  
  if (length(chunk_plots) > 0) {
    chunk_plots <- sort(chunk_plots)
    
    cat("\n\n## Biotype distribution plots by sample chunks\n\n")
    
    for (plot_file in chunk_plots) {
      plot_name <- basename(plot_file)
      part_label <- gsub("^biotype_distribution_percent_", "", plot_name)
      part_label <- gsub("\\.png$", "", part_label)
      part_label <- gsub("_", " ", part_label)
      
      cat(paste0("\n\n### ", part_label, "\n\n"))
      cat(paste0("![](", plot_file, ")\n\n"))
    }
  } else {
    cat("*No biotype distribution plots were found.*\n")
  }
} else {
  cat(paste0("*Missing plot:* `", full_biotype_plot, "`\n"))
}
```

```{r biotype-table}
show_table_if_exists(
  plot_path("biotype_qc", "biotype_distribution_counts_and_fraction.tsv"),
  "Biotype distribution table"
)
```

---

# 4. PCA on top variable genes

The PCA is calculated from the log-normalized expression matrix using the top variable genes.

## PCA scatter plot

```{r pca-scatter, fig.align='center', out.width='90%'}
include_png_if_exists(plot_path("pca", "pca_scatter_PC1_PC2.png"))
```

## Scree plot

```{r pca-scree, fig.align='center', out.width='80%'}
include_png_if_exists(plot_path("pca", "pca_scree.png"))
```
## PCA plots colored by sample metadata

```{r pca-metadata-plots, results='asis', echo=FALSE}
metadata_plot_dir <- plot_path("pca", "metadata_plots")

if (dir.exists(metadata_plot_dir)) {
  metadata_plots <- list.files(
    metadata_plot_dir,
    pattern = "^pca_scatter_PC1_PC2_by_.*\\.png$",
    full.names = TRUE
  )
  
  if (length(metadata_plots) > 0) {
    metadata_plots <- sort(metadata_plots)
    
    for (plot_file in metadata_plots) {
      plot_name <- basename(plot_file)
      plot_label <- gsub("^pca_scatter_PC1_PC2_by_", "", plot_name)
      plot_label <- gsub("\\.png$", "", plot_label)
      plot_label <- gsub("_", " ", plot_label)
      
      cat(paste0("\n\n### Colored by ", plot_label, "\n\n"))
      cat(paste0("![](", plot_file, ")\n\n"))
    }
  } else {
    cat("*No metadata-colored PCA plots were generated.*\n")
  }
} else {
  cat("*No metadata-colored PCA plots were generated.*\n")
}
```

## PCA variance explained

```{r pca-var-table}
show_table_if_exists(
  plot_path("pca", "pca_variance_explained.tsv"),
  "PCA variance explained"
)
```

---

# 5. Gene-set contamination scores

This section reports ssGSEA scores for platelet, erythroid/hemolysis, and endothelial marker sets.

## ssGSEA score versus PC correlation heatmap

```{r ssgsea-pc-heatmap, fig.align='center', out.width='100%'}
include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "ssgsea_score_pc_correlation_heatmap.png"))
```

## Platelet score vs PC1

```{r platelet-pc1, fig.align='center', out.width='80%'}
include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "platelet_score_vs_PC1_ssgsea.png"))
```

## Erythroid/hemolysis score vs PC1

```{r erythroid-pc1, fig.align='center', out.width='80%'}
include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "erythroid_score_vs_PC1_ssgsea.png"))
```

## Endothelial score vs PC1

```{r endothelial-pc1, fig.align='center', out.width='80%'}
include_png_if_exists(plot_path("gene_set_qc", "ssgsea", "endothelial_score_vs_PC1_ssgsea.png"))
```

## ssGSEA scores table

```{r ssgsea-score-table}
show_table_if_exists(
  plot_path("gene_set_qc", "ssgsea", "ssgsea_scores.tsv"),
  "ssGSEA scores"
)
```

## Marker matching statistics

```{r ssgsea-marker-stats}
show_table_if_exists(
  plot_path("gene_set_qc", "ssgsea", "ssgsea_marker_match_stats.tsv"),
  "Marker matching statistics"
)
```

---

# 6. Sex chromosome QC

This section visualizes XIST expression versus the mean expression of a small Y-linked gene panel.

```{r sex-scatter, fig.align='center', out.width='80%'}
include_png_if_exists(plot_path("sex_inference", "sex_inference_scatter.png"))
```

```{r sex-scores}
show_table_if_exists(
  plot_path("sex_inference", "sex_inference_scores.tsv"),
  "Sex chromosome QC scores"
)
```

```{r sex-marker-stats}
show_table_if_exists(
  plot_path("sex_inference", "sex_inference_marker_match_stats.tsv"),
  "Sex chromosome marker matching statistics"
)
```

---

# 7. HPC QC metric correlation with expression PCs

This section compares upstream HPC-derived QC metrics against expression PCs.

## HPC QC metric-PC correlation heatmap

```{r hpc-pc-heatmap, fig.align='center', out.width='100%'}
include_png_if_exists(plot_path("hpc_metric_pc_qc", "hpc_qc_pc_correlation_heatmap.png"))
```

```{r hpc-cor-table}
show_table_if_exists(
  plot_path("hpc_metric_pc_qc", "hpc_qc_pc_correlations.tsv"),
  "HPC QC metric-PC correlations"
)
```

## Z-scored HPC QC metrics

```{r hpc-z-table}
show_table_if_exists(
  plot_path("hpc_metric_pc_qc", "hpc_qc_metrics_zscore.tsv"),
  "Z-scored HPC QC metrics"
)
```

---

# 9. Notes

- Correlation heatmaps show **Pearson r** values.
- Significance stars, when present, use BH-adjusted p-values:
  - `*` adjusted p < 0.05
  - `**` adjusted p < 0.01
  - `***` adjusted p < 0.001
- Missing optional plots/tables usually mean that the corresponding optional input was not provided, or that the module was skipped.